Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-05-24 jss/EpiCompare-paper/EpiCompare-paper.tex
early comments on section 4 - more global thoughts

\textcolor{orange}{[[NEW TEXT}

@@

The Hagelloch data include a rich set of features at the individual
level. Recorded features include household members, school level,
household locations, date of first symptoms (prodromes), date of measles
rash, and even the alleged infector. A subset of the data is shown in
Table \ref{tab:hags-people}. With \pkg{EpiCompare}, we can easily
\textcolor{orange}{pre-process the data to} obtain the empirical
cumulative incidence function with respect to the measles rash
appearance (variable \code{ERU}) with the following tidy-style function,
\code{agents_to_aggregate()}. The function \code{agents_to_aggregate()}
is a key component of \pkg{EpiCompare}, allowing the user to easily
switch from an individual-level (i.e.~an agent) \sout{view} lens of a
disease to an aggregate \sout{level} lens. For example, the below code
shows how we can convert the agent data to a cumulative incidence
\textcolor{orange}{plot} of the measles
rash\sout{, in order to see how the disease spread through the population over time.}
We can then compare the cumulative incidence of the rash to the
cumulative incidence of the prodromes, i.e.~the initial symptoms. We do
this with the below code, and a part of the cumulative incidence data
output is shown in Table \ref{tab:cif-rash}. The argument
\code{integer_time_expansion} indicates whether we should include all
time points in the recorded range of the data or only when there is a
change in the incidence.

@@

One possible question of interest is the duration between initial onset
of prodromes and the appearance of the measles rash. Since
\code{agents_to_aggregate()} outputs a tidy-style data frame, it is a
simple task to plot the two sets of incidence curves on the same graph
(Fig. \ref{fig:cifs}).

@@

@@

The real power of \code{agents_to_aggregate()} lies in its ability to
aggregate over any number of pre-specified states. For example, the
Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the
time of infection and recovery, respectively of each individual. We can
then plot the SIR values through a time-invariant lens using
\pkg{ggplot2} and \pkg{ggtern} functions (as shown in Fig.
\ref{fig:hag-tern-raw}) or with our custom \code{geom},
\code{geom_aggregate}, which takes the raw agent data as input.

@@

Moreover, we can look at the outbreaks of the disease by group within
\code{agent_to_aggregate()} or \code{geom_aggregate()}. This allows us
to examine differences among the different groups of individuals. For
example, we show the time invariant outbreak by class level in Figure
\ref{fig:tern-class-data}. Immediately, we see that time invariant
infection curve is different for the pre-school class compared to the
1st class. In the 1st class, we see about 95\% of the class become
infected and less than 10\% of them having recovered, which may be
indicative of a super-spreading event. This suspicion is further
confirmed in that 26 of the 30 1st class students have been reportedly
infected by the same individual.

@@

Along with multiple epidemic states, the function
\code{agents_to_aggregate()} can also be extended to populations with
vital dynamics (e.g.~birth and death) and examples of this are shown in
the package vignette. In summary, \code{agents_to_aggregate()} is a
multi-purpose workhorse that may be leveraged to convert individual
level records into aggregate information that may be more useful for
some forms of epidemic modeling such as compartment modeling.

\hypertarget{modeling-and-simulation}{%
\subsection{Modeling and Simulation}\label{modeling-and-simulation}}

@@

Up to this point, we have used \pkg{EpiCompare} in the context of
observed data. We also want to compare statistical models, and
\pkg{EpiCompare} aids in that process via a simple yet flexible
individual-level
simulator\sout{, conversion tools for popular epidemic model packages, and model assessments.}
We demonstrate an example here.

We first try to model the Hagelloch data with a stochastic SIR model,
which we refer to as the `simple SIR.' In our full vignette
\textcolor{orange}{(available online),} we show how to fit this simple
SIR model via maximum likelihood and simulate from the model with those
best fit parameters. Our function \code{simulate_agents()} generates
individual level data according to discrete time multinomial draws,
which depend on the number of individuals in each state at the previous
time step and a matrix of transition probabilities. For example, the
below code generates 100 simulations of an outbreak of a disease with
one initial infector in a population of \(n= 188\) individuals.

@@

The result of our simulation is the object \code{agents} which is a
18800 \(\times\) 5 data frame, which details the time of entry into the
\(S\), \(I\), and \(R\) states for a given simulation. Before we examine
the results of this simple SIR model, we will also examine another, more
sophisticated SIR model, this time from the package \pkg{EpiModel}
\citep{Jenness2018}. Briefly, this model first fits a contact network to
the set of individuals, where the class of the child is a covariate. The
model then simulates a SIR-epidemic on that network.

@@

The output of this model is \code{epimodel_sir}, an object of class
\code{netsim}, which contains a plethora of modeling information.

@@

The next step is to compare the Simple SIR model to the EpiModel SIR
model. We provide the function \code{fortify_aggregate()}, which can
take objects from specialized classes of modeling output
\textcolor{orange}{(like those made by \code{netsim()})} and transform
it into a tidy-style data frame.

@@

We can then analyze the results of the two models side by side as
time-invariant epidemic curves. The results are shown in Figure
\ref{fig:hag-simple-sir}, where a 90\% prediction band is estimated from
the delta ball method for each of the two models. For the Simple SIR
model, we see that the data generally covers the data fairly well but
clearly misses the second peak of infection. We also see that the
prediction band is very large, covering up a large area of the ternary
plot. On the other hand, for the EpiModel network model, we see that the
prediction band covers the data quite well and takes up less area.

\begin{CodeChunk}
\begin{CodeInput}
R> both_models <- bind_rows(agg_model, fortified_net)
R> 
R> 
R> g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0),
+          aes(x = X0, y = X1, z = X2,
+               sim_group = sim, fill = Type),
+          alpha = .5,
+          conf_level = .90) 
\end{CodeInput}
\end{CodeChunk}

\begin{CodeChunk}
\begin{CodeInput}
R> g +   geom_path(data = both_models %>% filter(t !=0),
+             aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
+             alpha = .3, col = "gray40") + 
+     coord_tern() + theme_sir(base_size = 24) +
+   geom_point(data = hagelloch_sir,
+              aes(x = S, y = I, z =R), col = "black") +
+   labs(title = "Simple SIR model",
+        subtitle = "90% Prediction band and original data",
+        x = "S", y = "I", z = "R") +
+   scale_fill_manual(values = c("#006677", "#AA6600")) + 
+   facet_wrap(~Type) +
+   theme(legend.position = "bottom")
\end{CodeInput}
\begin{figure}[H]

@@

However, both models are not a good fit to the filamental path as
opposed to the individual points in \((S, I, R)\)-space. This can be
captured with the set of simulations both models predict (gray lines),
which all generally have a single defined peak of infection whereas the
data certainly looks like it has two distinct peaks, likely caused by
our assumed super-spreader event. This observation is backed up by the
below analysis that demonstrates that the estimated pseudo-density of
the observed epidemic (relative to the simulations from either model) is
much less likely then \textbf{any} of the simulations (reported in Table
\ref{tab:hags-extreme})\footnote{\textcolor{orange}{Ben, do we want to add another sentence or two explaining the two columns in the table?  The second one I think makes sense to me but not the first.}}
In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the
EpiModel network model is a better fit than the Simple SIR model, and 2)
the fit is only good at the
\sout{geometric filamental level as opposed to the epidemic trajectory filamental level.}
\textcolor{orange}{individual point level as opposed to the geometric filamental level.}

@@

Overall, \pkg{EpiCompare} aids in the data analysis pipeline for both
novice and expert practitioners and coders alike. These tools encourage
model and simulation exploration of many of the existing and
well-supported packages that already exist, and side-by-side comparison
thereof. Finally, we hope that practicioners will consider using
time-invariant analysis when trying to assess and compare epidemics and
epidemic models.

>>>>

\textcolor{orange}{[[NEWER TEXT}

\footnote{\textcolor{violet}{[Ben says: Generally I found this whole section to be pretty passive and not well motivated on why we'd actually do the analysis.]}}
To conclude our paper, we demonstrate the capabilities of
\pkg{EpiCompare} with a complete data analysis of a measles outbreak in
1861-1862 Germany. Specifically, we demonstrate how tools in
\pkg{EpiCompare} can be used in each step of the data analysis pipeline
\footnote{\textcolor{violet}{[Ben says: based on an earlier read of this section, I might suggest something like "and streamline the analysis process" - is that a selling point you want to highlight?]}}
\textcolor{violet}{\sout{(see Figure \ref{fig:pipeline})}}\footnote{\textcolor{violet}{[Ben says: It's unclear why you want to highlight the figure again - could you be clearere on that? / what does it add to the conv/ why should they reference it?]}}.
Additionally, we highlight how time-invariant analysis (see Section
\ref{sec:time-invariant}) can be used to enhance understanding of an
outbreak\footnote{\textcolor{violet}{[Ben says: I'm not sure this is necessary to state. Additionally - is it really true do we highlight this?]}}.

\footnote{\textcolor{violet}{[Ben says:The data introduction could be another subsection. Why does it fit better here?.   Other comment: This paragraph is pretty rambly. It is unclear what you want someone to take away from it. I might suggest highlight the fact that the data is "an ideal testing group for methodology".]}}
Before demonstrating \pkg{EpiCompare}, we provide some context for the
measles outbreak presented
here.\footnote{\textcolor{violet}{[Ben says: this first sentence is pretty sign-post-y yet it only relates to the next few sentences. Update so it's less sign-post-y.]}}
The data was originally organized by \cite{pfeilsticker1863}, later made
visible by \cite{oesterle1992}, and made available in an \proglang{R} by
\cite{surveillance2017}. This data set includes a rich collection of
features including household location, class level, and allged infector
ID, and is an ideal testing ground for methodology in infectious disease
epidemiology
\cite{Neal2004,britton2011,groendyke2012,becker2016}\footnote{\textcolor{violet}{[Ben says: This citation doesn't make sense - should it be "citep"?]}}.
In this data set, there are 188 children who became infected with the
measles over the course approximately 90 days.

\textcolor{orange}{[[LESS NEW TEXT}

@@

\footnote{\textcolor{violet}{[Ben says: The first two sentences is very similar to the data paragraph in the section above. Given it's not really connecting the 2 sections I suggest a rewrite - could move some of this stuff above.]}}The
Hagelloch data include a rich set of features at the individual level,
\textcolor{orange}{and the tools in \pkg{EpiCompare} help with pre-processsing and EDA}.
Recorded features include household members, school level, household
locations, date of first symptoms (prodromes), date of measles rash, and
even the alleged infector. A subset of the data is shown in Table
\ref{tab:hags-people}.
\textcolor{orange}{For example,}\footnote{\textcolor{violet}{[Ben says: I'm unclear of what this is actually an example of.]}}
with \pkg{EpiCompare}, we can easily
\textcolor{orange}{pre-process the data to} obtain the empirical
cumulative incidence function with respect to the measles rash
appearance (variable \code{ERU}) with the following tidy-style function,
\code{agents_to_aggregate()}. The function \code{agents_to_aggregate()}
is a key component of \pkg{EpiCompare}, allowing the user to easily
switch from an individual-level (i.e.~an agent) \sout{view} lens of a
disease to an aggregate \sout{level} lens. For
example\footnote{\textcolor{violet}{[Ben says: Same comment as before.]}},
the below code shows how we can convert the agent data to a cumulative
incidence \textcolor{orange}{plot} of the measles
rash\sout{, in order to see how the disease spread through the population over time.}
We can then compare the cumulative incidence of the rash to the
cumulative incidence of the prodromes, i.e.~the
initial\footnote{\textcolor{violet}{[Ben says: please be clearer on what these could be given you comparing them to start of the rash - which seems like an early symptom to me...]}}
symptoms\footnote{\textcolor{violet}{[Ben says: This action in the analysis pipeline is unmotivated - which naturally makes me want to ask "why would I do this?"]}}.
We do this with the below code, and a part of the cumulative incidence
data output is shown in Table \ref{tab:cif-rash}. The argument
\code{integer_time_expansion} indicates whether we should include all
time points in the recorded range of the data or only when there is a
change in the incidence.

@@

One possible question of interest is the duration between initial onset
of prodromes and the appearance of the measles
rash\footnote{\textcolor{violet}{[Ben says: You don't give a good definition of prodromes above, and you only use the name twice. Is this a super common term in Epi? I find it a bit taxing on the reader to remember what this is referring to.]}}.
Since \code{agents_to_aggregate()} outputs a tidy-style data frame, it
is a simple task to plot the two sets of incidence curves on the same
graph (Fig. \ref{fig:cifs}).

@@

\footnote{\textcolor{violet}{I'm confused why Figure 5 is included. What is the conclusion you'd like to take away? / Why do people create plots like this?}}

@@

The real power of \code{agents_to_aggregate()} lies in its ability to
aggregate over any number of pre-specified states. For example, the
Hagelloch data sets contains two columns, \code{tI} and \code{tR}, the
time of infection and recovery, respectively of each individual. We can
then\footnote{\textcolor{violet}{[Ben says: using "then" here captures a very progression step of analysis but I stopped here and asked "what is this following" - and the previous "step" occurred a paragraph back and wasn't described as a direct progression but just a possible thing to do.]}}
plot the SIR values through a time-invariant lens using \pkg{ggplot2}
and \pkg{ggtern} functions (as shown in Fig. \ref{fig:hag-tern-raw}) or
with our custom \code{geom}, \code{geom_aggregate}, which takes the raw
agent data as input.

@@

\footnote{\textcolor{violet}{I found this pargraph very unmotivated. I recommend first arguing why we might a care to look into the class subpopulation grouping. And maybe comment that is is a common desire for practioners.}}Moreover,
we can look at the outbreaks of the disease by group within
\code{agent_to_aggregate()} or \code{geom_aggregate()}. This allows us
to examine differences among the different groups of individuals. For
example, we show the time invariant outbreak by class level in Figure
\ref{fig:tern-class-data}. Immediately, we see that time invariant
infection curve is different for the pre-school class compared to the
1st class. In the 1st class, we see about 95\% of the class become
infected and less than 10\% of them having recovered, which may be
indicative of a super-spreading event. This suspicion is further
confirmed in that 26 of the 30 1st class students have been reportedly
infected by the same individual.

@@

\textcolor{violet}{\sout{Along with multiple epidemic states, the function}}
\code{agents_to_aggregate()}
\textcolor{violet}{\sout{can also be extended to populations with vital dynamics (e.g. birth and death) and examples of this are shown in the package vignette.  In summary, }}
\code{agents_to_aggregate()}
\textcolor{violet}{\sout{is a multi-purpose workhorse that may be leveraged to convert individual level records into aggregate information that may be more useful for some forms of epidemic modeling such as compartment modeling.}}\footnote{\textcolor{violet}{Is this not just a repeat of section 3?}}

\hypertarget{modeling-and-simulation}{%
\subsection{Modeling and simulation}\label{modeling-and-simulation}}

@@

Up to this point, we have used \pkg{EpiCompare} in the context of
observed data. \footnote{\textcolor{violet}{[Ben says: Why?]}}We also
want to compare statistical models, and \pkg{EpiCompare} aids in that
process via a simple yet flexible individual-level
simulator\sout{, conversion tools for popular epidemic model packages, and model assessments.}
We demonstrate an
example\footnote{\textcolor{violet}{[Ben says: to me this whole section is 1 example - as such this wording is confusing to me.]}}
here.

We first try to model the Hagelloch data with a stochastic SIR model,
which we refer to as the `simple
SIR.'\footnote{\textcolor{violet}{[Ben says: could / should this be thought of as a "base" model?]}}
In our full vignette \textcolor{orange}{(available online),} we show how
to fit this simple SIR model via maximum likelihood and simulate from
the model with those best fit
parameters\footnote{\textcolor{violet}{[Ben says: should you highlight that this is a common approach?]}}.
Our function
\code{simulate_agents()}\textbackslash footnote\{\textcolor{violet}{[Ben Should somethings be said about the }
\texttt{simulation\_SIR\_agents()} \textcolor{violet}{function?]}\}
generates individual level data according to discrete time multinomial
draws, which depend on the number of individuals in each state at the
previous time step and a matrix of transition probabilities. For
example, the below code generates 100 simulations of an outbreak of a
disease with one initial infector in a population of \(n= 188\)
individuals.

@@

The result of our simulation is the object \code{agents} which is a
18800 \(\times\) 5 data frame, which details the time of entry into the
\(S\), \(I\), and \(R\) states for a given simulation.
\footnote{\textcolor{violet}{[Ben says: please motivate - through model comparison?]}}Before
we examine the results of this simple SIR model, we will also examine
another, more sophisticated SIR model, this time from the package
\pkg{EpiModel} \citep{Jenness2018}. Briefly, this model first fits a
contact network to the set of individuals, where the class of the child
is a
covariate\footnote{\textcolor{violet}{[Ben says: Make this connect better to the problem at hand - why do you think you should build this bigger model?]}}.
The model then simulates a SIR-epidemic on that network.

@@

The output of this model is \code{epimodel_sir}, an object of class
\code{netsim}, which contains a plethora of modeling information.
\footnote{\textcolor{violet}{[Ben says: what's the point for this sentence. It also doesn't flow/ connect to previous and later text.]}}

@@

The next step is to compare the simple SIR model to the EpiModel SIR
model. We
provide\footnote{\textcolor{violet}{[Ben says: the phrase "We provide" is very passive / distance from the current demonstration at hand. Moreover section 3 already phrases things this way.]}}
the function \code{fortify_aggregate()}, which can take objects from
specialized classes of modeling output
\textcolor{orange}{(like those made by \code{netsim()})} and transform
it into a tidy-style data frame.

@@

We can then analyze the results of the two models side by side as
time-invariant\footnote{\textcolor{violet}{[Ben says: this isn't a clear phrrase here - what are you trying to say?]}}
epidemic curves. The results are shown in Figure
\ref{fig:hag-simple-sir}, where a 90\% prediction band is estimated from
the delta
ball\footnote{\textcolor{violet}{[Ben says: this has never been discussed anyway.]}}
method for each of the two models. For the Simple SIR model, we see that
the data generally covers the data fairly well but clearly misses the
second peak of
infection\footnote{\textcolor{violet}{[Ben says: This could be better movitated with talk of model fit...]}}.
We also see that the prediction band is very large, covering up a large
area of the ternary plot. On the other hand, for the EpiModel network
model, we see that the prediction band covers the data quite well and
takes up less area.

\begin{CodeChunk}
\begin{CodeInput}
R> both_models <- bind_rows(agg_model, fortified_net)
R> 
R> 
R> g <- ggplot() + geom_prediction_band(data = both_models %>% filter(t != 0) %>%
+                   mutate(Type = factor(Type, levels = c("Simple SIR", 
+                                                         "EpiModel SIR"))),
+          aes(x = X0, y = X1, z = X2,
+               sim_group = sim, fill = Type),
+          alpha = .5,
+          conf_level = .90) 
\end{CodeInput}
\end{CodeChunk}

\textcolor{violet}{[Ben says: In figure 8 I changed the order of the facets given we talk about the simple model first and its more like the "base" model. I think the title should be changed?]}

\begin{CodeChunk}
\begin{CodeInput}
R> g +   geom_path(data = both_models %>% filter(t != 0) %>%
+                   mutate(Type = factor(Type, levels = c("Simple SIR", 
+                                                         "EpiModel SIR"))),
+             aes(x = X0, y = X1, z = X2, group = paste(Type, sim)),
+             alpha = .3, col = "gray40") + 
+     coord_tern() + theme_sir(base_size = 24) +
+   geom_point(data = hagelloch_sir,
+              aes(x = S, y = I, z =R), col = "black") +
+   labs(title = "Simple SIR model",
+        subtitle = "90% Prediction band and original data",
+        x = "S", y = "I", z = "R") +
+   scale_fill_manual(values = c("#006677", "#AA6600")) + 
+   facet_wrap(~Type) +
+   theme(legend.position = "bottom")
\end{CodeInput}
\begin{figure}[H]

@@

However, both models are not a good fit to the filamental path as
opposed to the individual points in \((S, I, R)\)-space. This can
be\footnote{\textcolor{violet}{[Ben says: this is a very passive way to say such things. Try being more direct.]}}
captured with the set of simulations both models predict (gray lines),
which all generally have a single defined peak of infection whereas the
data certainly looks like it has two distinct peaks, likely caused by
our assumed super-spreader event. This observation is backed
up\footnote{\textcolor{violet}{[Ben says: describe this?]}} by the below
analysis that demonstrates that the estimated pseudo-density of the
observed epidemic (relative to the simulations from either model) is
much less likely then \textbf{any} of the simulations (reported in Table
\ref{tab:hags-extreme})\footnote{\textcolor{orange}{Ben, do we want to add another sentence or two explaining the two columns in the table?  The second one I think makes sense to me but not the first.}}
In conclusion, \pkg{EpiCompare} makes it clear that, at a glance, 1) the
EpiModel network model is a better fit than the Simple SIR model, and 2)
the fit is only good at the
\sout{geometric filamental level as opposed to the epidemic trajectory filamental level.}
\textcolor{orange}{individual point level as opposed to the geometric filamental level.}\footnote{\textcolor{violet}{[Ben says: how would this look with the time plots? Do we add value here?]}}

@@

\footnote{\textcolor{violet}{I think this paragraph captures some good goals, but I don't think we've done some of this. For example - we don't really highlight novice/expert usage, and we don't highlight side-by-side comparisons of models.}}Overall,
\pkg{EpiCompare} aids in the data analysis pipeline for both novice and
expert practitioners and coders alike. These tools encourage model and
simulation exploration of many of the existing and well-supported
packages that already exist, and side-by-side comparison thereof.
Finally, we hope that practicioners will consider using time-invariant
analysis when trying to assess and compare epidemics and epidemic
models.
